---
title: "Maze Part 2"
author: "Andrew Marx"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Maze Part 2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

required <- c("viridisLite")
if (!all(sapply(required, requireNamespace, quietly = TRUE))) {
  knitr::opts_chunk$set(eval = FALSE)
}

library(samc)
library(raster)
library(gdistance)
library(viridisLite)

maze = matrix(
  c(1,1,1,1,0,1,0,1,0,1,0,1,0,0,1,1,1,0,0,1,
    0,1,0,1,1,1,0,1,1,1,1,1,1,0,1,0,1,1,0,1,
    0,1,0,0,1,0,0,0,0,0,1,0,0,0,0,0,1,0,0,1,
    0,1,1,0,1,1,0,1,1,1,1,0,1,0,1,0,1,1,1,1,
    0,0,0,0,1,0,0,1,0,0,1,1,1,1,1,0,0,0,0,1,
    1,1,1,1,1,0,1,1,0,1,1,0,1,0,1,0,1,1,0,1,
    1,0,0,0,1,0,0,1,0,0,1,0,0,0,1,0,0,1,0,1,
    1,1,1,0,1,1,0,1,1,0,1,1,0,1,1,1,0,1,1,1,
    1,0,1,0,1,0,0,0,1,0,0,0,0,1,0,0,0,0,1,0,
    1,0,0,0,1,1,1,0,1,1,1,1,0,1,1,0,1,0,1,1,
    1,1,1,0,1,0,1,1,1,0,1,0,0,0,1,1,1,0,1,0,
    0,1,0,0,0,0,0,0,1,0,1,1,0,1,1,0,0,0,1,1,
    1,1,1,1,0,1,1,0,0,0,0,1,0,0,1,0,1,0,1,0,
    0,0,0,1,0,0,1,1,1,0,1,1,0,1,1,0,1,1,1,1,
    0,1,0,0,0,1,1,0,1,0,0,0,0,1,0,0,0,0,0,1,
    0,1,1,1,1,1,0,0,0,0,1,1,1,1,0,1,1,1,0,1,
    0,0,0,0,0,1,0,1,1,1,1,0,1,0,0,0,1,0,0,1,
    1,1,0,1,0,1,1,1,0,0,0,0,0,0,1,0,1,1,1,1,
    0,1,1,1,0,0,0,1,1,1,0,1,0,1,1,1,1,0,1,0,
    1,1,0,1,1,1,1,1,0,1,1,1,1,1,0,1,0,0,1,1),
  nrow = 20
)

maze <- raster(maze, xmn = 0.5, xmx = ncol(maze) + 0.5, ymn = 0.5, ymx = nrow(maze) + 0.5)
maze[maze==0] <- NA

#
# Get info about the shortest path through the maze using gdistance
#
points <- xyFromCell(maze, c(1, 400))

lcd <- (function() {
  tr <- transition(maze, function(x) 1/mean(x), 4)
  tr <- geoCorrection(tr)

  list(dist = costDistance(tr, points),
       path = shortestPath(tr, points[1, ], points[2, ], output="SpatialLines"))
})()

# Setup a simple color palette
vir_col <- viridis(3)

# End of maze
maze_end <- maze * 0
maze_end[20, 20] <- 1

tr <- list(fun = function(x) 1/mean(x), dir = 4, sym = TRUE)

samc_obj <- samc(maze, maze_end, tr_args = tr)

start <- locate(samc_obj, data.frame(x = 1, y = 20))
finish <- locate(samc_obj, data.frame(x = 20, y = 1))
```


## Introduction

In [Part 1](example-maze-part1.html) of this series, we explored the applications of absorbing Markov chains in the context of a simple perfect maze. In Part 2, we will expand on our simple maze is various ways and explore how these changes affect our interpretations when using the different metrics offered in the samc package.

## Setup

This example reuses the [libraries](example-maze-part1.html#libraries1), [setup](example-maze-part1.html#setup1), and original [samc object](example-maze-part1.html#create-the-samc-object-1) code from Part 1.


## Fidelity

We'll start by incorporating fidelity into the samc object. There's potentially many different ways that fidelity could be applied to the maze, but we'll keep things simple by using it to create a delay in movement at intersections. The motivation might be hesitancy or decision making when the individual is presented with the choice of three or more paths. For simplicity, we'll only set the fidelity at all intersections to the same `0.1` value, which basically means that once an individual is at an intersection, there is a 10% probability they will stay in the intersection from one time step to the next.
```{r, fig.show='hold', fig.width=7, fig.height=5, fig.align='center'}
# Intersections determined using a moving window function
maze_ints <- focal(maze, w = matrix(c(NA, 1, NA, 1, 1, 1, NA, 1, NA), nrow = 3, ncol = 3), fun = function(x){sum(!is.na(x)) > 3}, pad = TRUE)
maze_ints[is.na(maze)] <- NA
maze_ints <- maze_ints * 0.1

plot(maze_ints, main = "Intersections", col = vir_col[c(2, 3)], axes = FALSE, box = FALSE, asp = 1)
plot(rasterToPolygons(maze), border = 'black', lwd = 1, add = TRUE)
points(points, pch = c('S', 'F'), cex = 1, font = 2)
```

Fidelity actually changes the $P$ matrix underling the samc object, which means we have to create a new samc object:
```{r}
samc_ints <- samc(maze, maze_end, maze_ints, tr_args = tr)
```

To start, let's see how the new fidelity input affects our expected time to finish:
```{r}
# Our original results from Part 1
survival(samc_obj)[start]
cond_passage(samc_obj, start, finish)

# Our results with fidelity at intersections
survival(samc_ints)[start]
cond_passage(samc_ints, start, finish)
```

Intuitively, since we've added "hesitation" to the movement with no other changes, the expected time to finish increases. We also see that incorporating fidelity in this particular example does not affect the relationship between `survival()` and `cond_passage()`. We can also see that changing the fidelity for this example does not change the probability of visiting any particular cell:
```{r}
all.equal(dispersal(samc_obj, origin = start), dispersal(samc_ints, origin = start))
```

It does, however, change the number of times each cell is expected to be visited:

```{r, fig.show='hold', fig.width=7, fig.height=5, fig.align='center'}
visit_orig <- visitation(samc_obj, origin = start)
visit_ints <- visitation(samc_ints, origin = start)

all.equal(visit_orig, visit_ints)

# Let's plot the difference to see if there is a noticeable pattern
visit_diff <- map(samc_obj, visit_ints) - map(samc_obj, visit_orig)

plot(visit_diff, col = viridis(256), main = "Visits Per Cell (Difference)", axes = FALSE, box = FALSE, asp = 1)
plot(rasterToPolygons(maze_end), border = 'black', lwd = 1, add = TRUE)
points(points, pch = c('S', 'F'), cex = 1, font = 2)
```

It looks like with fidelity present, the intersections are clearly seeing significant number of visits. Basically, since a "visit" represents a transition to a cell from one time step to the next, the presence of fidelity means that the metric is counting transitions from a cell to itself as well. Interestingly, the legend in this figure seems to indicate that the increase for the intersections might actually be 10%, or the same as the fidelity probabilities. It also seems like the non-intersections experienced no change. Let's check these ideas.
```{r, fig.show='hold', fig.width=7, fig.height=5, fig.align='center'}
# First, let's see which cells changed.
# Unfortunately, this is another case where float precision is problematic.
# We can't just compare to 0, so we assume that anything within a very small
# tolerance is equivalent to 0
tolerance = sqrt(.Machine$double.eps) # Default tolerance in functions like all.equal()
print(tolerance)

plot(visit_diff > tolerance, col = vir_col[c(2, 3)], main = "Visits With Non-Zero Difference", axes = FALSE, box = FALSE, asp = 1)
plot(rasterToPolygons(maze_end), border = 'black', lwd = 1, add = TRUE)
points(points, pch = c('S', 'F'), cex = 1, font = 2)

# Second, let's see what the percent change is for our non-zero differences.
visit_perc <- (visit_ints - visit_orig) / visit_orig
visit_perc[visit_perc>tolerance]
```

It turns out that there is no change in the number of expected visits for non-intersections. It also turns out that our hunch for the intersections was only partially true; the change is constant, but it's 1/9 instead of 0.1 or 10%.

The most interesting change we might see from incorporating fidelity will actually be with the `distribution()` metric. Recall from Part 1 that there was an alternating pattern with the cells when changing the time steps. With fidelity, this effect still exists, but not to the same degree:
```{r, fig.show='hold', fig.width=7, fig.height=5, fig.align='center'}
dist_ints <- distribution(samc_ints, origin = start, time = 20)

plot(map(samc_ints, dist_ints), col = viridis(256), main = "Location at t=20", axes = FALSE, box = FALSE, asp = 1)
plot(rasterToPolygons(maze), border = 'black', lwd = 1, add = TRUE)
points(points, pch = c('S', 'F'), cex = 1, font = 2)


dist_ints <- distribution(samc_ints, origin = start, time = 21)

plot(map(samc_ints, dist_ints), col = viridis(256), main = "Location at t=21", axes = FALSE, box = FALSE, asp = 1)
plot(rasterToPolygons(maze), border = 'black', lwd = 1, add = TRUE)
points(points, pch = c('S', 'F'), cex = 1, font = 2)
```

Given a sufficient amount of time, the cumulative effect of having fidelity present will almost entirely eliminate the alternating pattern. Even at time step 200, the alternating pattern is essentially gone, at least visually:
```{r, fig.show='hold', fig.width=7, fig.height=5, fig.align='center'}
dist_ints <- distribution(samc_ints, origin = start, time = 200)

plot(map(samc_ints, dist_ints), col = viridis(256), main = "Location at t=200", axes = FALSE, box = FALSE, asp = 1)
plot(rasterToPolygons(maze), border = 'black', lwd = 1, add = TRUE)
points(points, pch = c('S', 'F'), cex = 1, font = 2)


dist_ints <- distribution(samc_ints, origin = start, time = 201)

plot(map(samc_ints, dist_ints), col = viridis(256), main = "Location at t=201", axes = FALSE, box = FALSE, asp = 1)
plot(rasterToPolygons(maze), border = 'black', lwd = 1, add = TRUE)
points(points, pch = c('S', 'F'), cex = 1, font = 2)
```

For comparison, here's the original samc object using the same time steps:
```{r, fig.show='hold', fig.width=7, fig.height=5, fig.align='center'}
dist <- distribution(samc_obj, origin = start, time = 200)

plot(map(samc_obj, dist), col = viridis(256), main = "Location at t=200", axes = FALSE, box = FALSE, asp = 1)
plot(rasterToPolygons(maze), border = 'black', lwd = 1, add = TRUE)
points(points, pch = c('S', 'F'), cex = 1, font = 2)


dist <- distribution(samc_obj, origin = start, time = 201)

plot(map(samc_obj, dist), col = viridis(256), main = "Location at t=201", axes = FALSE, box = FALSE, asp = 1)
plot(rasterToPolygons(maze), border = 'black', lwd = 1, add = TRUE)
points(points, pch = c('S', 'F'), cex = 1, font = 2)
```

## Dead End Avoidance

Technically, the package doesn't offer the ability to "look ahead" at future states to adjust the transient probabilities. In other words, if a route would eventually lead to a dead end, there's nothing in the `samc()` function or the metric functions to account for that or model the possibility that an individual in the maze can see down a hallway. We can, however, fake it somewhat by adjusting the resistance map so that the dead ends have a much higher resistance. This will reduce the probability of an individual entering into the dead end, almost as if they looked ahead. We can use a sliding window function to create this new map:
```{r, fig.show='hold', fig.width=7, fig.height=5, fig.align='center'}
# Dead ends
maze_ends <- focal(maze, w = matrix(c(NA, 1, NA, 1, 1, 1, NA, 1, NA), nrow = 3, ncol = 3), fun = function(x){sum(!is.na(x)) == 2}, pad = TRUE)
maze_ends[is.na(maze)] <- NA
maze_ends <- maze_ends * 9 + 1
maze_ends[20, 20] <- 1

plot(maze_ends, main = "Dead Ends", col = vir_col[c(2, 3)], axes = FALSE, box = FALSE, asp = 1)
plot(rasterToPolygons(maze), border='black', lwd=1, add=T)
points(points, pch = c('S', 'F'), cex = 1, font = 2)
```

We have given the dead ends a resistance value of 10, and kept everything else at 1. This is fairly high and means that dead ends will only rarely be entered. Since we've modified our resistance map, we need to update the samc object. We will not use our fidelity data from the previous section, which will allow us to compare directly against the model created in Part 1.
```{r}
samc_ends <- samc(maze_ends, maze_end, tr_args = tr)
```

Hypothetically, since an individual can now "look ahead" (remember, we faked it), they should be able to get through the maze faster because they are spending far less time in dead ends. We can see this easily:
```{r}
# Our original results from Part 1
survival(samc_obj)[start]
cond_passage(samc_obj, start, finish)

# Our results with dead ends
survival(samc_ends)[start]
cond_passage(samc_ends, start, finish)
```
Since the dead ends have a lower probability of being transitioned to, the `dispersal()` and `visitation()` metrics should obviously reflect that:
```{r, fig.show='hold', fig.width=7, fig.height=5, fig.align='center'}
disp_ends <- dispersal(samc_ends, origin = start)

plot(map(samc_obj, disp_ends), col = viridis(256), main = "Probability of Visit", axes = FALSE, box = FALSE, asp = 1)
plot(rasterToPolygons(maze), border = 'black', lwd = 1, add = TRUE)
points(points, pch = c('S', 'F'), cex = 1, font = 2)


visit_ends <- visitation(samc_ends, origin = start)

plot(map(samc_obj, visit_ends), col = viridis(256), main = "Visits Per Cell", axes = FALSE, box = FALSE, asp = 1)
plot(rasterToPolygons(maze), border = 'black', lwd = 1, add = TRUE)
points(points, pch = c('S', 'F'), cex = 1, font = 2)
```

The effect is more obvious with `visitation()` and the expected number of visits; the probability illustration is more subtle compared to the original results in Part 1. This could be explored similarly to how some of the differences in the fidelity section are illustrated, an exercise that will be left to interested readers.


## Traps

It's fairly to trivial to take our example and add in lethal traps to our maze; it just requires updating our absorption input to the `samc()` function. The key thing to keep in mind is that `samc()` expects the total absorption, so we will only be giving it a single absorption input. However, our metrics do allow us to tease apart the role that different sources of absorption will have in the model. There are two different approaches to setting this up:

1. We start with a single total absorption input. Then we take that input and decompose it into multiple absorption components.
1. We start with multiple absorption components. Then we take those inputs and combine them into a single total absorption input.

The choice really depends on the data available and the goals of the project. For example, the former strategy is useful if we've somehow measured total absorption for a model and want to explore different hypotheses for how it breaks down into different types of absorption. The former is useful if we already have direct knowledge of different sources of absorption. For this example, we're going with the latter; we already have one absorption component for our finish point, and now we are going to create a second simple one that represents a few traps with a 0.2 or 20% absorption probability:
```{r, fig.show='hold', fig.width=7, fig.height=5, fig.align='center'}
# Traps absorption layer
maze_traps <- maze * 0
maze_traps[17, 3] <- 0.2
maze_traps[1, 6] <- 0.2
maze_traps[6, 20] <- 0.2

plot(maze_traps, main = "Traps", col = vir_col[c(2, 3)], axes = FALSE, box = FALSE, asp = 1)
plot(rasterToPolygons(maze), border='black', lwd=1, add=T)
points(points, pch = c('S', 'F'), cex = 1, font = 2)
```

In order to use this, we will have to create a new samc object:
```{r}
abs_total <- maze_end + maze_traps

samc_traps <- samc(maze, abs_total, tr_args = tr)
```

We're again only modifying the basic example from Part 1 for easy comparison. Like our other changes, let's see how long we now expect it to take for an individual to finish the maze:
```{r}
# Our original results from Part 1
survival(samc_obj)[start]
cond_passage(samc_obj, start, finish)

# Our results with fidelity at intersections
survival(samc_traps)[start]
cond_passage(samc_traps, start, finish)
```

The results are drastically different from what we've seen before. First, we no longer have the clear relationship between `survival()` and `cond_passage()`. `survival()` no longer gives us information on how long it will take to finish; instead it now tells us long it will take to either finish or die to one of our traps. This also drastically changes what we see when plotting the results of `survival()` (note the change in figure title from Part 1):
```{r, fig.show='hold', fig.width=7, fig.height=5, fig.align='center'}
survive_traps <- survival(samc_traps)

# Note the updated title from part 1
plot(map(samc_obj, survive_traps), col = viridis(256), main = "Expected Time to Absorption", axes = F, box = F, asp = 1)
plot(rasterToPolygons(maze), border = 'black', lwd = 1, add = TRUE)
points(points, pch = c('S', 'F'), cex = 1, font = 2)
```

We also see drastically different results from Part 1 when looking at visitation probability and number of visits:
```{r, fig.show='hold', fig.width=7, fig.height=5, fig.align='center'}
disp_traps <- dispersal(samc_traps, origin = start)

plot(map(samc_traps, disp_traps), col = viridis(256), main = "Probability of Visit", axes = FALSE, box = FALSE, asp = 1)
plot(rasterToPolygons(maze), border = 'black', lwd = 1, add = TRUE)
points(points, pch = c('S', 'F'), cex = 1, font = 2)


visit_traps <- visitation(samc_traps, origin = start)

plot(map(samc_traps, visit_traps), col = viridis(256), main = "Visits Per Cell", axes = FALSE, box = FALSE, asp = 1)
plot(rasterToPolygons(maze), border = 'black', lwd = 1, add = TRUE)
points(points, pch = c('S', 'F'), cex = 1, font = 2)
```

Importantly, we can longer use the same technique as Part 1 where we used visitation probabilities of 1 to identify the route through the maze; it only works in very specialized cases. The reason is simple: since the individual can now be absorbed in other locations, there is a non-zero probability that they never reach the finish. However, we can still use the same technique to see something interesting:
```{r, fig.show='hold', fig.width=7, fig.height=5, fig.align='center'}
# Ideally, we would just use `as.numeric(disp == 1)`, but float precision means
# that what we think is `1` isn't always `1` with computers, so it won't always
# work. We work around it by subtracting 1 and seeing if the result fits within
# a very small tolerance
tolerance = sqrt(.Machine$double.eps) # Default tolerance in functions like all.equal()
print(tolerance)
disp_traps_route <- as.numeric(abs(disp_traps - 1) < tolerance)

plot(map(samc_traps, disp_traps_route), col = vir_col[c(2, 3)], main = "dispersal() == 1", axes = FALSE, box = FALSE, asp = 1)
plot(rasterToPolygons(maze_end), border = 'black', lwd = 1, add = TRUE)
points(points, pch = c('S', 'F'), cex = 1, font = 2)
```

It shows us part of the solution we saw before, but only up to the first maze intersection that leads to two or more possible sources of absorption.

The inclusion of multiple absorption states now allows us to explore metrics that were not useful of Part 1. Starting with `mortality()`, we can visualize where we would expect the individual to be absorbed:
```{r, fig.show='hold', fig.width=7, fig.height=5, fig.align='center'}
mort_traps <- mortality(samc_traps, origin = start)

plot(map(samc_traps, mort_traps), col = viridis(256), main = "Absorption Probability", axes = FALSE, box = FALSE, asp = 1)
plot(rasterToPolygons(maze), border = 'black', lwd = 1, add = TRUE)
points(points, pch = c('S', 'F'), cex = 1, font = 2)
```

That quite possibly is not what we would have expected. Why does the finish point look like it's 0? Let's look at the numbers and see if we can figure it out from there:
```{r}
mort_traps[mort_traps > 0]

mort_traps[finish]
```

There's only a 2.1% chance of our individual finishing the maze! But if we stop to think about it, it actually makes sense. Recall the "Probability of visiting a cell" and "Visits per cell" sections from Part 1; the individual spends most of their time in the early part of the maze. That means they have a lot more exposure to our first trap and, consequently, are more likely to be absorbed there. And if we look at the trap farthest from the start, reaching it requires barely missing the finish, so consequently it has only a 0.28% of being the source of absorption, or an order of magnitude lower probability than just finishing.

The package allows us to break things down so that we can investigate the role of different types of mortality more easily. Remember, we started with two absorption layers that we summed together; now that the samc object has been created, we can provide it that information:
```{r}
# Naming our rasters will make things easier and less error-prone later
names(maze_end) <- "Finish"
names(maze_traps) <- "Traps"

samc_traps$abs_states <- raster::stack(maze_end, maze_traps)
```

By doing so, the `mortality()` metric now returns a list with information about not just the total absorption, but the individual components we specified. This allows us to individually access and visualize the role of different types of absorption:
```{r, fig.show='hold', fig.width=7, fig.height=5, fig.align='center'}
mort_traps <- mortality(samc_traps, origin = start)

str(mort_traps)

plot(map(samc_traps, mort_traps$Finish), main = "Absorption Probability (Finish)", col = viridis(256), axes = FALSE, box = FALSE, asp = 1, nc = 1)
plot(rasterToPolygons(maze), border = 'black', lwd = 1, add = TRUE)
points(points, pch = c('S', 'F'), cex = 1, font = 2)

plot(map(samc_traps, mort_traps$Traps), main = "Absorption Probability (Traps)", col = viridis(256), axes = FALSE, box = FALSE, asp = 1, nc = 1)
plot(rasterToPolygons(maze), border = 'black', lwd = 1, add = TRUE)
points(points, pch = c('S', 'F'), cex = 1, font = 2)
```

We can also now make use of the `absorption()` metric:
```{r}
absorption(samc_traps, origin = start)
```

The output from this is quite simple: it tells us the probability that the individual will experience a particular type of absorption. So as we saw before, there is a 2.1% chance of finishing the maze. But now we see that there is a 97.9% total probability that absorption will occur in one of the three traps. This is different from the `mortality()` metric, which tells us the absorption probabilities at each cell. The advantage of this example is that it's easy to see the relationship with our original `mortality()` results above; once you start mixing multiple types of absorption at the same cell or transient state, it becomes more difficult to see these relationships.
