---
title: "Multiple Absorption Tutorial"
author: "Andrew Marx"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple Absorption Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

required <- c("viridisLite")
if (!all(sapply(required, requireNamespace, quietly = TRUE))) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

## Introduction

Starting in v1.4.0, the samc package included support for multiple sources of absorption when creating an samc object. Additionally, when multiple sources of absorption are provided, the results of the mortality() metrix are now decomposed into its individual and total components, which are returned in a named list. When only a single absorption input is provided, the function will work exactly like it did previously to maintain backwards compatability with existing code.

*Note: only the asymptotic calculations are currently decomposed. The short-term mortality metrics still need to be updated.*


## Setup

For this example, we will use the example data included in the package, but will semi-randomly divide/decompose the absorption matrix into two separate layers. These layers are used for illustrative purposes and do not represent anything meaningful.

```{r, message = FALSE, fig.show='hold'}
# First step is to load the libraries. Not all of these libraries are stricly
# needed; some are used for convenience and visualization for this tutorial.
library("raster")
library("samc")
library("viridisLite")

# "Load" the data. In this case we are using data built into the package.
# In practice, users will likely load raster data using the raster() function
# from the raster package.
res_data <- samc::ex_res_data
abs_data <- samc::ex_abs_data
occ_data <- samc::ex_occ_data

# Generate some random values that will be used as proportions for dividing the
# absorption data.
p1 <- runif(length(abs_data), max = 0.3)
p2 <- 1 - p1

# Divide our absorption data into two layers. When added together, these two
# layers should be identical to the original
abs_data_a <- abs_data * p1
abs_data_b <- abs_data * p2

all.equal(abs_data_a + abs_data_b, abs_data)
```


## Total (Single) Absorption

The use of multiple absorption layers is mainly of interest for the mortality() metric, which calculates the probability of absorption at different transient states or locations. By providing multiple absorption layers, we can tease apart the role that different sources of absorption might have in a model. To start, we will work with the total absorption.

```{r, fig.width = 6.5, out.width = '100%', fig.height= 3, fig.align = "center"}
# Setup the details for our transition function
tr <- list(fun = function(x) 1/mean(x), # Function for calculating transition probabilities
           dir = 8, # Directions of the transitions. Either 4 or 8.
           sym = TRUE) # Is the function symmetric?

# Create the samc object using the original "total" absorption layer
samc_obj <- samc(res_data, abs_data, tr_args = tr)

# Let's say we're interested in where absorption is expected to occur throughout
# the model if starting from a single location
mort <- mortality(samc_obj, origin = 1)

# When only working with total absorption, this version of mortality() returns
# a named vector
str(mort)

# Let's visualize it
mort_map <- map(samc_obj, mort)
plot(mort_map, xlab = "x", ylab = "y", col = viridis(256))
```


## Multiple Absorption

Since our data is a matrix object, we will provide the decomposed absorption layers in a list object. When using raster data, it would be provided as either a RasterStack or a RasterBrick. The layers in the list/RasterStack/RasterBrick can be named, which will carryover into the results, which can help with code readability when accessing different parts of the results.

```{r, fig.width = 6.5, out.width = '100%', fig.height= 3, fig.align = "center"}
# Let's attach absorption layers to our samc object
samc_obj$abs_states <- list(abs1 = abs_data_a,
                            abs2 = abs_data_b)

# Now rerun the analysis
mort_multiple <- mortality(samc_obj, origin = 1)

# Let's note the differences from what was returned before. Here, we have a list
# of named vectors. The first is for the total absorption. The last two are for
# the subdivided or decomposed absorption inputs
str(mort_multiple)

# Let's visualize it
multiple_map <- map(samc_obj, mort_multiple)
multiple_map <- raster::stack(multiple_map) # Convert the list to a RasterStack for plotting
plot(multiple_map, xlab = "x", ylab = "y", col = viridis(256), nc = 1, nr = 3)


# Let's check some things.

#First, the results of the decomposed layers in the list should add up to the total result
all.equal(multiple_map[[1]], multiple_map[[2]] + multiple_map[[3]])
# Alternatively, we could use the layer names:
all.equal(multiple_map$total, multiple_map$abs1 + multiple_map$abs2)

# Second, notice in the plots above that the result for the single input and the total
# result for the multiple input look very similar? That's because they are identical
all.equal(mort, mort_multiple$total)


# To remove the additional layers
samc_obj$abs_states <- NA
```

